Variational Monte Carlo on bosonic systems

### 2018-03-23 project_1/article/article.tex
Filled in initial non-interacting system text

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

		Scanning the entire space of variational parameters can be a somewhat despairing task. To aid us in the search we
		therefore implement the method of gradient descent\footnote{Also known as steepest descent, easily confused
		with the steepest descent method for integation.}. In any optimum we generally have that the derivative should equal
		zero. In our case, we want to find the minimum of the expected local energy with respect to some variational
		parameter $\alpha$, so that
		\begin{equation}
			\nabla_\alpha \expval{E_L(\mathbf{R}; \alpha)} = 0.
		\end{equation}	
		The foundation for  the method of gradient descent lies in the fact that some analytic
		\footnote{We use the term analytic quite recklessly.} function $F(\alpha)$ decreases fastest in the direction
		of $-\nabla_\alpha F(\alpha)$. This means that we could eventually get to an ensuing minimum by trailing along the
		pathway of the following recursive relation,
		\begin{equation}
			\alpha_{k+1} = \alpha_k - \gamma \nabla_{alpha k} F(\alpha_k).
		\end{equation}
		If $\gamma$ is small enough we should have $f(\alpha_k) \geq F(\alpha_{k+1})$ for all $k$ and hopefully we will
		experience convergence to the desired minimum.

		For our special case : INSERT DERIVATION OF LOCAL ENERGY??

@@

    We will start without importance sampling, that is by employing the simpler and more naïve "brute force" Monte Carlo
    sampling. Moreover, we will compute the value for the local energy analytically as well as with a numerical approach.
    In the numerical approach we employ the well-know approximation for the second derivative,
    \begin{equation}
    		f''(x)= \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
    \end{equation}

@@

	For the first

@@

@@

@@

         \begin{table}
             \caption{BLEH}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&181.69371&0.25533&0.98756&199.95366\\
                     0.3&142.61591&0.10782&0.97849&195.80928\\
                     0.4&129.88615&0.04051&0.96862&190.44553\\
                     0.5&127.29926&0.00595&0.95749&195.10368\\
                     0.6&129.97630&0.03300&0.94595&193.22755\\
                     0.7&135.67382&0.05785&0.93330&191.96391\\
                     0.8&143.23238&0.07286&0.92061&192.72399\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:50_interacting}
         \end{table}

         \begin{table}
             \caption{BLEH}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&375.04401&0.51475&0.98527&745.21438\\
                     0.3&296.76879&0.20214&0.97534&741.04425\\
                     0.4&270.60830&0.08113&0.94522&723.06823\\
                     0.5&266.37263&0.02020&0.95314&702.71000\\
                     0.6&272.51171&0.07366&0.94112&689.12862\\
                     0.7&285.10285&0.11381&0.92895&707.48248\\
                     0.8&301.40609&0.15458&0.91648&707.86118\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:100_interacting}
         \end{table}

         \begin{figure}
             \includegraphics[width=250px]{../data/figures/problem_e.pdf}
             \caption{BLEH}
             \label{fig:interacting}
         \end{figure}

>>>>

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

		Scanning the entire space of variational parameters can be a somewhat despairing task. To aid us in the search we
		therefore implement the method of gradient descent\footnote{Also known as steepest descent, easily confused
		with the steepest descent method for integation.}. In any optimum we generally have that the derivative should equal
		zero. In our case, we want to find the minimum of the expected local energy with respect to some variational
		parameter $\alpha$, so that
		\begin{equation}
			\nabla_\alpha \expval{E_L(\mathbf{R}; \alpha)} = 0.
		\end{equation}	
		The foundation for  the method of gradient descent lies in the fact that some analytic
		\footnote{We use the term analytic quite recklessly.} function $F(\alpha)$ decreases fastest in the direction
		of $-\nabla_\alpha F(\alpha)$. This means that we could eventually get to an ensuing minimum by trailing along the
		pathway of the following recursive relation,
		\begin{equation}
			\alpha_{k+1} = \alpha_k - \gamma \nabla_{\alpha k} F(\alpha_k).
		\end{equation}
		If $\gamma$ is small enough we should have $f(\alpha_k) \geq F(\alpha_{k+1})$ for all $k$ and hopefully we will
		experience convergence to the desired minimum.

		For our special case : INSERT DERIVATION OF LOCAL ENERGY derivative wrt alpha??

@@

    We will start without importance sampling, that is by employing the simpler and more naïve "brute force" Monte Carlo
    sampling. Moreover, we will compute the value for the local energy analytically as well as with a numerical approach.
    In the numerical approach we employ the well-know approximation for the second derivative,
    \begin{equation}
    		f''(x) \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
    \end{equation}

@@

@@

	The first initial observation of note is that we have found an optimum for the system in terms of a minimum
	energy at $\alpha=\frac{1}{2}$. Moreover, notice how the standard deviation disappears completely.
	This should already be apparent from \autoref{eq:closed_form_natural_units_local_energy}, as the term
	involving particle positions disappears for the optimal $\alpha$.

@@

	Systems of optimum variational parameter warrants further investigation, as summed up in 
	\autoref{tab:initial_optimum_comparison}. The analytical results are quite predictable and 
	monotonous; the expected local energy $\expval{E_l}$ is perfectly proportional by a factor
	$\frac{3}{2}$ to the number of particles and the standard deviation $\sigma_b$ is zero for
	systems of any number of particles. The numerical results tells a more tortuous tale. As the 
	number of particles increase, the uncertainty also increases rapidly. For the largest system,
	with $N = 500$ particles, the energy misses its target critically - the true energy is more than
	one standard deviation away. Lastly, the cpu time is notably higher for the numeric scheme.

	\subsubsection{Introducing Importance Sampling}

	Next we will introduce importance sampling..

@@

         \begin{table}
             \caption{$50$ interacting particles.}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&181.69371&0.25533&0.98756&199.95366\\
                     0.3&142.61591&0.10782&0.97849&195.80928\\
                     0.4&129.88615&0.04051&0.96862&190.44553\\
                     0.5&127.29926&0.00595&0.95749&195.10368\\
                     0.6&129.97630&0.03300&0.94595&193.22755\\
                     0.7&135.67382&0.05785&0.93330&191.96391\\
                     0.8&143.23238&0.07286&0.92061&192.72399\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:50_interacting}
         \end{table}

         \begin{table}
             \caption{$100$ Interacting particles.}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&375.04401&0.51475&0.98527&745.21438\\
                     0.3&296.76879&0.20214&0.97534&741.04425\\
                     0.4&270.60830&0.08113&0.94522&723.06823\\
                     0.5&266.37263&0.02020&0.95314&702.71000\\
                     0.6&272.51171&0.07366&0.94112&689.12862\\
                     0.7&285.10285&0.11381&0.92895&707.48248\\
                     0.8&301.40609&0.15458&0.91648&707.86118\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:100_interacting}
         \end{table}

         \begin{figure}
             \includegraphics[width=250px]{../data/figures/problem_e.pdf}
             \caption{Local energies as a function of variational parameter $\alpha$
             				 for different number of particles.}
             \label{fig:interacting}
         \end{figure}
